---
title: "Assessment 3"
format: html
editor: visual
---

### Read Data

```{r}
library(tidyverse)
bor_nestlings <- read_csv(here::here("data/73_species.csv"))
bor_traits <- read_csv(here::here("data/Traits_73_species.csv"))
avonet <- readxl::read_excel(here::here("data/AVONET%20Supplementary%20dataset%201.xlsx"), sheet = "AVONET1_BirdLife")
```

### Explore the data

Are nestlings showing up earlier in the year over time?

```{r}
all_birds_trend <- bor_nestlings %>% 
  group_by(Year) %>% 
  summarize(mean_doy = mean(Dayofyear))

ggplot(all_birds_trend, aes(Year, mean_doy)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

That combined data for all 73 species, grouped by year. Let's also group by species to get a species breakdown by year.

```{r}
species_trends <- bor_nestlings %>% 
  group_by(Year, Species) %>% 
  summarize(mean_doy = mean(Dayofyear),
            .groups = "drop")

ggplot(species_trends, aes(Year, mean_doy, color = Species)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

Weeeeelll that's a pukey plot if I've ever seen one...Let's just look at the 5 most data rich species.

```{r}
#How many data points for each unique species are represented in our dataset? 
data_richness <- bor_nestlings %>% 
  count(Species)

#Arrange in descending order (most to least data points) 
most_rich <- data_richness %>% 
  arrange(desc(n)) %>% 
  slice(1:5)

#Okay now lets looks at a less pukey species trends for just the 5 most data rich spp.
most_rich_trends <- bor_nestlings %>% 
  filter(Species %in% most_rich$Species) %>% #oooh indexing by name AND position, FANCY!
  group_by(Species, Year) %>% 
  summarize(mean_doy = mean(Dayofyear), 
            .groups = "drop")

#TA-DA! A less pukey plot! Hooray!
ggplot(most_rich_trends, aes(Year, mean_doy, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Cool! Most species have a similar trend to one another except for one which has an opposite trend. Interesting...But which species have the strongest trends?

```{r}
# Finds the slope of the relationship between y and x
trend <- function(x, y) {
  xy_lm <- lm(y ~ x)
  coef(xy_lm)[2]
}

# Calculate the trend for all species
bor_trends <- species_trends %>% 
  group_by(Species) %>% 
  summarize(doy_trend = trend(Year, mean_doy))
```

Quick spot check of 2 species

```{r}
soi <- c("ARDCIN", "LARMIN")
species_trends %>% 
  filter(Species %in% soi) %>% 
  ggplot(aes(Year, mean_doy, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

### MY Turn!

Find the species with the most positive and most negative trends, then plot just those.

```{r}
nrow_bor_trends <- nrow(bor_trends) # Use this later
bor_extreme <- bor_trends %>% 
  arrange(doy_trend) %>% # Sort by the day of year trend
  slice(c(1, nrow_bor_trends)) # Keep just the first (most negative trend) and last (most positive trend) rows

# Now plot the most positive and negative species trends
species_trends %>% 
  filter(Species %in% bor_extreme$Species) %>% 
  ggplot(aes(Year, mean_doy, color = Species)) + 
    geom_point() +
    geom_smooth(method = "lm")
```

### Let's Bring in More Data

Next we\'re going to test the hypothesis that smaller birds have more flexible phenology, i.e. the absolute value of the trend is correlated with smaller body sizes.

To test our hypothesis, we need to add AVONET data to our phenology data by joining.

```{r}
# First, select and rename important columns
avonet_size <- select(avonet, SciName = Species1, Mass_g = Mass)
bor_sciname <- select(bor_traits, 
                      Species = Abbreviation, 
                      SciName = `Scientific name`)
```

Now join them all together.

```{r}
bor_trends_size <- bor_trends %>% 
  left_join(bor_sciname, by = "Species") %>% 
  left_join(avonet_size, by = "SciName") %>% 
  mutate(abs_trend = abs(doy_trend))

# Plot it
ggplot(bor_trends_size, aes(Mass_g, abs_trend)) +
  geom_point() +
  geom_smooth(method = "lm")
```

That\'s kind of hard to interpret. Make a new variable, `bor_trends_size2`, that removes the missing values in `Mass_g` and keeps only birds smaller than 2000 g with an absolute trend less than 1.

```{r}
bor_trends_size2 <- bor_trends_size %>% 
  drop_na(Mass_g) %>% 
  filter(Mass_g < 2000,
         doy_trend < 1)
  
  
ggplot(bor_trends_size2, aes(Mass_g, abs_trend)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Our hypothesis is not supported! We don't have evidence to say that smaller birds have more flexible phenology.
